Full factorial construction of synthetic microbial communities

Version 1

Abstract

Abstract

Constructing combinatorially complete species assemblages is often necessary to dissect the complexity of microbial interactions and to find optimal microbial consortia. At the moment, this is accomplished through either painstaking, labor intensive liquid handling procedures, or through the use of state-of-the-art microfluidic devices. Here we present a simple, rapid, low-cost, and highly accessible liquid handling methodology for assembling all possible combinations of a library of microbial strains, which can be implemented with basic laboratory equipment. To demonstrate the usefulness of this methodology, we construct a combinatorially complete set of consortia from a library of eight Pseudomonas aeruginosa strains, and empirically measure the community-function landscape of biomass productivity, identify the highest yield community, and dissect the interactions that lead to its optimal function. This easy to implement, inexpensive methodology will make the assembly of combinatorially complete microbial consortia easily accessible for all laboratories.

Introduction

Microbial consortia are playing a growingly important role in biotechnology. They possess substantial potential advantages over monocultures, including larger metabolic capabilities, division of labor, and a potentially higher ecological and evolutionary stability. 1 Synthetic microbial communities are being engineered for goals as diverse as degrading pollutants, 2 producing high-value molecules such as biofuels, 3,4 vitamins, 5,6 and flavonoids, 7,8 and preventing the invasion of pathogens. 9

To fully realize the promise of microbial consortia in biotechnology we must develop tools for their optimization. Given a library of candidate strains, which consortia should we form if we wish to maximize a function of interest? 10 One way to address this question is purely empirical, following a full factorial design in which we form every possible assemblage and simply select whichever one is best. 11–15 Even for candidate libraries of small size (less than 10 species) the combinatorial nature of this strategy makes the assembly of all possible communities painstaking and tedious, limiting the scale and replicability of full factorial design. The number of unique liquid handling events required to form all possible combinations of m species scales as m2m−1, as each of the m species needs to be added to each of the 2m−1 consortia where it is present. Manually making all combinations is therefore draining, slow, and prone to human error. Because of the long time required to complete such a task, the risk of contamination is high, and replication is constrained. Perhaps for this reason, studies reporting full factorial construction of microbial consortia are very rare, 11,13,14,16 and the field has largely relied on fractional factorial design, where only a subset of representative species combinations are constructed. 10,17–19

The ability to form full factorial combinations of a library of species would have benefits beyond biotechnology. In ecology, forming every possible species assemblage is required if we wish to investigate the complex, combinatorial nature of microbial interactions 16,20–25 and the role that high-order interactions play in the function and dynamics of microbial communities. 19,24,26–30 In sum, the quantitative understanding of high-order microbial interactions, the empirical mapping of community assembly networks, the quantitative exploration of community-function landscapes, and the development of predictive models of microbial community assembly and function all require us to be able to reliably assemble communities of growing complexity. 15,18,30

Robotic liquid handlers can facilitate the task of assembling a full combinatorial set of microbial communities. 11 While this alleviates some of the challenges associated to combinatorial liquid handling, particularly those associated with human error, others remain: the execution of hundreds or thousands of pipetting events is still inevitably slow, and as a result additional equipment (i.e., a HEPA filter) is required to avoid contamination. Robotic liquid handlers are also expensive, technically sophisticated equipment and they are not routinely available. As an alternative to deal with the combinatorial complexity of community assembly, Blainey et al. have introduced a droplet-based microfluidic system (kChip) that is capable of forming hundreds of thou-sands of species assemblages, 27,31 facilitating their full factorial construction. Notwithstanding their power and unparalleled high-throughput, microdroplet-based approaches such as kChip and others 32 are still state-of-the-art experimental methods, requiring specialized equipment and training that are not yet available to the vast majority of research groups worldwide.

In this paper, we describe a simple, rapid, inexpensive, and highly accessible liquid handling methodology for the full factorial design of microbial consortia and inocula, which requires only basic laboratory equipment and which can be easily implemented by the vast majority of research laboratories. Using a standard multichannel pipette, a single user can manually assemble all possible combinations of up to 10 species in less than one hour, a timescale that is shorter than the replication time of most bacteria in minimal media. Higher-dimensional designs are possible too, with the only limitation being the plasticware that is required. In addition to providing detailed protocols for a specific set of conditions (e.g., volume of the culture, density of the bacteria, etc.) we provide an R script to help the user tailor the protocol to their needs. To illustrate the usefulness of this approach, we apply it to empirically characterize the full, combinatorially complete community-function landscape of a model microbial consortium consisting of eight Pseudomonas aeruginosa strains. We characterize the full absorption spectrum of all consortia and, as a model function, we determine their total biomass. We show that the implementation of our protocol, enabling the full factorial assembly of all strain combinations, allows us to quantitatively determine the relationship between community diversity and function, identify optimal strain combinations, and characterize all pairwise and higher-order interactions among all members of the consortia. We expect this methodology will rapidly expand the number of factorially constructed microbial consortia in the literature.

Methods

In what follows, we describe the reasoning behind our protocol in full detail. Readers interested in the practical implementation of this logic are referred to the “Full protocol description” section of the Supplementary Methods, where we provide a step-by-step guide for the assembly of all consortia from a particular example, consisting of a library of eight microbial isolates. In addition, we provide an R script which can be used to generate similar step-by-step guides for different numbers of species and pipetting volumes.

Logic of the assembly protocol

The mathematical basis of our method lies in identifying each microbial consortia by a unique binary number. For a set of m species, we can represent any consortium (which we generically call c) as where xk = 0, 1 represents the absence (0) or presence (1) of species k in the consortium (Figure 1A). For example, for m = 6 species, the consortium where only species 1, 2, and 3 are present would be represented as 000111.

Logic of the assembly protocol.

(A) We consider a set of m species (in this illustration, m = 8), and we identify each of the 2m consortia that can be formed with them by the binary number c = xmxm−1· · · x2x1 (with xk = 1, 0 representing the presence/absence of species k, respectively; see equation 1). (B) Our protocol is based on the idea that one could assemble each of the 23 = 8 combinations of species 1 to 3 in the first column of a 96-well plate. Iteratively duplicating these consortia and adding species 4 to m to them would eventually result in all 2m combinations of species, as described in the main text. (C) Each of the m species would be present in the indicated wells (see equation 2). (D) Positioning of the 2m consortia within the 96-well plates used in the protocol.

Binary numbers are the natural way to denote combina-stance, tions in mathematics, and indeed the correspondence between the two sets is often used in combinatorics. For inof m species. The same number counts how many binary strings of length m contain n ones (equivalently, n zeros).

Now, the most important aspect of this notation is that merging two disjoint consortia becomes a simple addition: if we combine consortium 110000 with consortium 000011, the resulting consortium is represented by the binary number 110011, which is the sum of the binary representations of the initial consortia: 110000 + 000011 = 110011. Note, however, that this operation only works for disjoint consortia. If we add two consortia sharing one or more species, binary addition is no longer a good representation: combining consortium 100001 with consortium 000001 should result in the first consortium again, 100001 (we are for now ignoring potential differences in species abundances/densities, but will address this issue later in the text). Yet, the addition of the two binary representations leads to 100010, which does not represent the appropriate consortium (remember that in binary addition 01 + 01 = 10). As we will see in what follows, our protocol minimizes the number of liquid handling events by making extensive use of this addition property, but only for disjoint consortia.

The second property that we will use to our advantage is that 96-well plates have 8 rows, which is a power of 2 (23 = 8). This allows us to form all combinations from a 3-species set (say, species 1, 2, and 3) in one column of the plate. It is convenient to arrange these 8 consortia following the order of their binary representation: the empty consortium (000) in the first well, followed by 001, 010, 011, 100, 101, 110, and 111; corresponding to the decimal numbers from 0 to 7 in increasing order.

We can now duplicate these eight consortia, pipetting them into the second column of the 96-well plate, and then pipette species 4 (1000) into all wells of this second column using a multichannel pipette (Figure 1B). This operation is equivalent to the binary addition of consortium 1000 (species 4 alone) with each of the starting eight consortia, and results in all 24 = 16 consortia which can be assembled with species 1 to 4: in the first column, as explained before, we have 0000, 0001, 0010, 0011, 0100, 0101, 0110, and 0111 (decimal numbers 0 to 7); while in the second column we now have 1000, 1001, 1010, 1011, 1100, 1101, 1110, and 1111 (decimal numbers 8 to 15, Figure 1B). We can next duplicate each of these 16 consortia, pipetting the first and second columns of the plate into the third and fourth columns, respectively. If we then add species 5 (10000) to each of the duplicated consortia, we generate all 25 = 32 combinations of species 1 to 5 (Figure 1B).

In what follows, we will describe how this logic can be implemented most efficiently in practice, but hopefully the reader can already intuit how to generalize this algorithm for consortia with more species. Note that for more than 6 species, multiple 96-well plates will be required: for instance, with m = 7 one could form 27 = 128 potential consortia, more than what a single 96-well plate can fit. Although we will not make use of them here, the algorithm is also generalizable for 384-well plates, as these have 16 rows. In this case, we would use the first column to assemble all consortia made up by species 1 to 4, and then proceed as described above for all subsequent species.

Plate arrangement

Following the logic we just discussed, it is straightforward to notice that species 4 will be present in all wells of columns 2, 4, 6, 8… of the 96-well plate. In turn, species 5 will be present in alternating pairs of columns: 3, 4, 7, 8, 11, 12… Species 6 will be present in alternating sets of 4 columns: 5, 6, 7, 8, 13, 14, 15, 16… (Figure 1C). Note that 96-well plates only have 12 columns: what we here call columns 13 to 24 would in practice be columns 1 to 12 of a second 96-well plate, columns 25 to 36 would correspond to a third plate, and so on. By a similar reasoning, species 1 to 3 will be present in alternating sets of 1, 2, and 4 rows (instead of columns) respectively (Figure 1C). In summary, we have that species k will be present in:

Species k present in: Conveniently, this arrangement makes it so consortia are located in order from top to bottom and from left to right within the plate(s): the first column contains the consortia corresponding to the binary representations of decimal integers 0 (00000) to 7 (00111), the second column corresponds to decimal integers 8 (01000) to 15 (01111), the third column to decimal integers 16 (10000) to 23 (10111), and so on.

Homogenizing species’ densities across consortia

As we just saw, it would be possible to form all combinations of m species simply by pipetting each one of them rowwise (for species 1 to 3) or column-wise (for species 4 to m) in their corresponding positions (equation 2) with the aid of multichannel pipettes. There is, however, one important caveat: if we pipette equal volumes of each species’ monocultures, we will end up with variable volumes across wells. As an example, if we pipette a volume v0 of each monoculture, we will have a volume of 2v0 in well D01 of the 96-well plate (where consortium 011 is located), but a volume of 3v0 in well H01 (where consortium 111 is located). Species 1 is present in both of these example consortia at the same total population size, but at different population densities differing by a factor of 2/3.

Often, we may want the density of each species to be consistent across consortia. To achieve this, we can pipette additional liquid (sterile water, growth medium, saline buffer…) into each well in order to compensate for the differences in volume. Note that, prior to this density homogenization step, the maximum volume in a well would be mv0, corresponding to the consortium with all m species. This naturally also corresponds to the consortium where each of the individual species is present at the lowest population density. In every other consortium (say, consortium c), the volume of buffer (which we denote vB (c)) that we would need to add in order to reach this minimum population density is given by where H (c) represents the Hamming weight of the binary representation of consortium c. The Hamming weight of a binary number is the number of ones in its representation, for instance, H (0000) = 0, H (0010) = 1, H (1111) = 4, etc. Thus, H (c) simply counts the number of species present in consortium c.

In practice, we do not need to manually pipette the buffer well-by-well. The convenient arrangement of the plate allows us to streamline this process. In what follows, we use i and j as the row and column indexes of the plate (from top to bottom and from left to right, respectively). For instance, well C02 (containing consortium 1010) would correspond to i = 3 and j = 2. As a reminder, we are considering that j = 13 to 24 correspond to columns 1-12 of a second 96-well plate, j = 25 to 36 correspond to a third plate, etc.

In order to streamline the pipetting of buffer, it would be convenient to find an expression for H (c) in terms of the positional indexes i and j. Now, due to the particular arrangement of consortia in the plate, we can divide the binary representation of a consortium c into two. First, the right-most three bits of c are determined by the row in which it is located. For instance, all consortia in row 7 will always have a binary representation ending in 110, independently of the column in which they appear. In general, the three rightmost digits of c will be equal to B (i − 1), the binary representation of I − 1 with i being the index of the row where consortium c is located. Similarly, the leftmost m − 3 bits of c will depend only on the column, in the following way: take a column j and ignore the three rightmost digits; then the binary representation of all consortia in that column will be the same, and equal to B (j − 1). For instance, with m = 8, the binary representation of all consortia in column 4 is 00011x3x2x1 — the last three bits may be 0 or 1 depending on the specific well, but the first five bits are common to all wells in the column. Therefore, in order to calculate the Hamming weight of a given consortium, we just need to sum the Hamming weight of the row and the Hamming weight of the column: We can use equation 4 to rewrite equation 3 in terms of the positional indexes i and j. After some algebra, we arrive at While terms in this equation could be arranged in a more compact form, equation 5 as it is allows us to implement the buffer pipetting in a very efficient way: We can start by pipetting liquid buffer column-wise (a volume v0 [m − 3 − H (B (j −1))] into every well of column j). We can then pipette the buffer row-wise, pipetting a volume v0 [3 −H (B (i− 1))] into every well of row i. Furthermore, the row-wise buffer pipetting can be bypassed if we homogenize the densities of the starting consortia formed with species 1 to 3, as we explain in detail in the Supplementary Methods (section “Full protocol description for m = 8 species”).

Results

Proof of concept: construction of colorant combinations

As a first demonstration of the method, and to establish its feasibility and accuracy, we used commercial food colorants and temperas to build all 256 combinations made up from eight colors. We diluted these colorants in water such that all of them had comparable maximum absorbances in the range from 380 nm to 780 nm. Each color exhibits a different absorbance spectrum in this range (Figure 2A).

Full factorial construction of synthetic colorant combinations.

(A) We consider a set of eight synthetic colorants, each of which exhibits a different absorbance spectrum in the range 380-780 nm. (B) As an example, we show the empirically measured absorbance spectrum (solid black line) for the combination of colorants 1, 3, and 8 (10000101). The dashed black line represents the additive expectation, i.e., the sum of the spectra of the three constituent colorants (colored dashed lines). (C) Relative deviation between the empirical spectrum and the additive expectation (see equation 6) for the example combination 10000101 (left) and for all mixtures of 2 or more colorants (right). Relative errors were only computed when the true absorbance was above 0.1 A.U..

Using synthetic colorants to test our protocol has various advantages. First, color differences are clearly visible to the naked eye, making it easy to keep track of the location of each colorant. Second, we do not expect these colorants to interact with one another, or to do so very weakly. This means that the absorbance spectrum of any combination of colorants may be reasonably expected to match the sum of the individual spectra of each colorant. Therefore, the magnitude of the deviations between an empirical spectrum and this additive expectation (Figure 2B) can be used to estimate an upper bound for the pipetting error of the protocol. We can obtain an estimation of the accuracy of our protocol by examining these deviations.

We followed the protocol described above, setting m = 8 and v0 = 25 μL (a detailed step-by-step protocol for these parameters is given in the Supplementary File protocol 8species 25uL.txt), to assemble all 256 color combinations. The absorbance spectrum of each combination was quantified using a MultiSkan SkyHigh plate reader (Thermo Scientific). The implementation of the protocol required three 96-well plates, 8 Falcon tubes, an 8-channel pipette and a 12-channel pipette. Completing the whole protocol took a single experimentalist less than an hour. Figure S1 shows the three 96-well plates at six different steps of the protocol.

Figure 2B shows the absorbance spectrum for an example colorant mixture (10000101), comparing it to the additive expectation that its three constituent colorants (1, 3, and 8) do not interact. We estimated the pipetting error of the protocol as the relative deviation (denoted δ) between the empirical spectra and the additive expectations at all wavelengths (Figure 2B), that is: where we have denoted Absλ (c) the absorbance at a wavelength λ (between 380 and 780 nm) of the colorant mixture c, and (c) the additive expectation (i.e., the sum of the absorbances of each individual constituent colorant).

In Figure 2C we show the distribution of relative deviations for the example mixture 10000101 at all wavelengths (left), as well as the distribution corresponding to all mixtures at all wavelengths (right). Figure S2 shows the ab-solute deviations for all mixtures at all wavelengths (median absolute deviation ∼ 0.025 A.U.). We found that relative deviations were generally small (Figure 2C: mean ∼ 5.8 %, median ∼ 4.9 %). We also found that the magnitude of these deviations did not increase with the number of colorants in a mixture (Figure S3), suggesting that the precision of our protocol is not compromised by the size of the set from which mixtures are assembled. Deviations were also roughly similar in magnitude across all wavelengths (Figure S4).

Full factorial design of synthetic bacterial consortia

To illustrate the potential applications of our protocol for finding optimal microbial consortia and quantitatively studying the complexity of microbial interactions, we adopted a collection of eight Pseudomonas aeruginosa strains obtained from a previous experiment. 33–35 Monocultures of each of these strains grown in liquid LB medium exhibit differences in their absorbance spectra (Figure 3A). Unlike the synthetic colorants, bacterial strains can be expected to interact with one another through a variety of potential mechanisms, including competition for limiting resources or facilitation which may lead to changes in the population sizes of different species. 15,19

Co-culturing combinations of bacterial isolates.

(A) We consider eight P. aeruginosa strains obtained from a previous experiment. 33–35 Here we represent the absorbance spectra of the eight monocultures of these strains. Solid lines and shaded areas represent means and standard deviations across three independent biological replicates. (B) We followed our protocol to assemble different combinations of strains (in this example, the consortium formed by strains 2, 3, and 8, represented as 10000110). Each consortium was diluted into fresh medium by a factor of 1:400 and allowed to re-grow for 40 h, after which the absorbance spectrum of the culture was measured. Solid black line: empirical spectrum of consortium 10000110 (mean of three independent biological replicates, shaded region represents the standard deviation across them). Dashed black line: additive expectation if the three strains did not interact. Dashed colored lines: monoculture spectra of the three constituent strains.

We first cultured each strain separately by resuspending a single colony into 15 mL of LB medium in 50 mL conical Falcon tubes. Strains were allowed to grow to carrying capac-ity for 24 h at 37 °C with no agitation. Then, monocultures were fully homogenized and used to assemble all 256 strain combinations following the same protocol described before (detailed steps can be found in the Supplementary Methods, section “Full protocol description for m = 8 species”). Phosphate buffer saline (1 × PBS) was used to adjust volumes/densities as indicated in the protocol. After completing our protocol (which took a single experimentalist less than one hour), each consortium was inoculated into fresh LB medium (0.5 μL of inoculum into 200 μL of medium for a dilution factor of 1:400; note that our protocol design makes it so this is equivalent to inoculating 0.5 μL/8 = 0.0625 μL of each strain’s monoculture into the fresh medium) in 96-well plates. Samples were incubated still at 37 °C for 40 h, after which the absorbance spectrum of each sample was quantified in the same plate reader we used before.

Figure 3B shows the absorbance spectrum for one example consortium after the 40 h incubation period. Consistent with our reasoning, this spectrum deviates strongly from the additive expectation (sum of each monoculture spectrum), indicating a prominent effect of interactions between strains. Figure 4 shows the spectra for all consortia with two or more members, where deviations from the additive expectation are similarly large.

Full factorial design of synthetic microbial consortia.

We represent the absorbance spectra of all consortia (of 2 or more strains) that can be formed with our library of eight P. aeruginosa isolates. As in Figure 3B, solid black lines and shaded areas represent the mean and standard deviation of the co-culture spectrum across three biological replicates; the dashed colored lines represent the individual monoculture spectra; and dashed black lines represent the sum of these individual spectra.

Because we have constructed all possible assemblages, we can systematically investigate the interactions between species and how they may result in different community-level properties and functions. As an illustration, we focus on the absorbance at 600 nm (Abs600), a metric of total biomass in our cultures. We found a non-monotonic relationship between community biomass and community richness (number of strains), peaking at 3 strains (Figure 5A). Our protocol allows us not only to characterize this type of broad-scale relationships between diversity and function, but also to identify specific high-performing consortia (Figure 5B). In our experiment, the consortium with the highest biomass contains only three strains (consortium 00001101, containing strains 1, 3, and 4). In Figure 5C we represent the full map between the composition and function (biomass) of all 256 consortia in our experiment (i.e., the community structure-function landscape 15,16,19,22).

Full factorial design of microbial consortia enables the dissection of microbial interactions.

We focus on the absorbance at 600 nm as a proxy for the biomass of a consortium. (A) Consortium function (biomass quantified as Abs600) against the number of strains in the consortium. Dots and error bars represent the average and standard deviation across all consortia with a same number of strains. (B) Top three consortia exhibiting the highest biomass across all 28 combinations. (C) We represent the function (biomass; values shown are averages across three biological replicates) for the consortia assembled using our protocol. Nodes correspond to the different consortia, edges connect consortia that differ in the presence of a single particular strain. (D) Detail showing the effect of including strain 1, strain 2, and both, in a background consortium (in this example, the background corresponds to consortium 11001000). Red: the functional effect of strain 1 is quantified as the difference in function between the background consortium and the consortium resulting from the inclusion of that strain. Green: the functional interaction between strains 1 and 2 in this particular background is quantified as the difference between the additive expectation (dashed line) and the empirical value for the function of the consortium containing both strains, as explained in the main text. (E) Detail of the functional landscape corresponding to the three strains that are part of the overall highest-biomass consortium (00001101). Inset: functional interactions within this 3-strain consortium. Order 1 corresponds to the monoculture functions. Order 2 corresponds to pairwise species-by-species interactions as defined in panel D of this figure. Order 3 corresponds to the third-order interaction between these strains, defined as the deviation between the empirical function of the trio with respect to the null expectation from the additive and pairwise interactions. 19,29 19,29 (F) Distributions of functional effects of the eight strains used in our experiments across all of the 128 backgrounds where each may be included. (G) Distributions of functional interactions between all pairs of strains across all backgrounds where both may be included. (H) Fraction of functional variance due to interactions of different orders (order 1 corresponds to the additive effects of each strain). The fractions were computed as explained in the Supplementary Methods. 30 Dashed line is the cumulative sum. (I) We have recently shown that the functional effect of a strain is often predictable from the function of the background consortium through simple, linear regressions. 10 For the eight P. aeruginosa strains in our experiment, linear regressions do an excellent job at linking these functional effects with the function of the background consortia.

From this type of data, one can readily quantify the sign and magnitude of strain interactions with respect to the community-level function of interest, in our case total biomass. In Figure 5D we show that the inclusion of strain 1 in a given consortium (which we refer to as an ecological background) results in an increase in biomass, and the same is true for the inclusion of strain 2. We call the difference in function between two consortia with and without a given focal strain as the functional effect of that strain in that ecological background. 10 If two strains did not interact, we may expect that including both of them in a same ecological background would result in a change in function equal to the sum of their two individual functional effects (see Figure 5D, dashed line represents this additive expectation). If the empirical function deviates from the additive expectation, the two strains engage in a functional interaction (denoted as ϵ in Figure 5D) in that particular ecological background. 10,19

Following this definition, one could dissect the interaction structure of all eight P. aeruginosa strains in our system. As an illustration, in Figure 5E we show the community-function landscape corresponding to all consortia that could be formed solely with strains 1, 3, and 4, i.e., the trio that maximizes biomass yield (see Figure 5B-C indicating that the maximum Abs600 is achieved by consortium 00001101). Although pairwise interactions are negative within this consortium, the three strains engage in a positive third-order functional interaction (see inset in Figure 5E) which rescues the function of the trio. This observation highlights the potential of full factorial design to interrogate and generate hy-potheses with respect to the mechanisms that govern the emergence of community-level functions, particularly (but not only) in high-performing consortia.

Beyond this particular consortium, in Figure 5F we represent the distributions of functional effects of all eight strains in our library across all of the 128 ecological backgrounds where they may be included. We found that the functional effects of all eight P. aeruginosa strains exhibit large variation in sign and magnitude, indicating that the contribution of a strain to the total community biomass largely depends on the presence or absence of additional community members in the ecological background. This indicates a prominent role of inter-strain interactions in this system.

In Figure 5G we represent the magnitude of the functional interactions between all pairs of strains i and j (denoted ϵij) across all the ecological backgrounds where the two may be included. We found that these pairwise interactions are highly variable across backgrounds. Pairwise interactions being sensitive to the presence or absence of additional community members can be interpreted as a signature of higher-order ecological interactions (HOIs). 19,36,37 By this definition, HOIs do exist in our particular experimental system (as evidenced by the variation in the pairwise interaction terms ϵij shown in Figure 5G); however, most of the variance in community biomass can be explained by low-order interactions (Figure 5H, the fraction of functional variance attributed to pairwise and higher-order interactions was computed as described elsewhere; 30 see Supplementary Methods and Figure S5 for details). This has also been shown to be the case for other datasets of microbial community function. 30

We have recently shown that the aggregate effect of pairwise and higher-order microbial interactions often results in emergent statistical patterns, which make it so the functional effect of a strain may be predictable from the function of the ecological background where that strain is included. 10 The relationships between the functional effect of a strain and the function of the background community can often be captured through simple linear models, which mirror the patterns of global epistasis commonly reported in genetics. This phenomenon is also observed in the particular set of P. aeruginosa strains that we analyzed here (Figure 5I, R2 averages 0.54), each of which exhibits a different relationship between its functional effect and the background function. Adopting the same analytical approach we used in previous work, 10 we find that such strong global epistasis patterns deviate from the null expectation of a random community-function landscape (Figure S6).

Together, these analyses illustrate how a full factorial construction of microbial consortia may allow us both to identify optimal assemblages and to dissect how the interplay of pairwise and higher-order interactions shape community-level properties and functions. To our knowledge, the experiments we have just reported represent one of barely a handful of datasets consisting of every potential combination from a library of strains. The fact that it was done in under an hour per replicate gives a sense of the potential our design has to dramatically expand the number of empirical community-function landscapes in the literature.

Discussion

We have introduced a rapid, simple, and inexpensive protocol for assembling all possible combinations of a given set of species using basic laboratory equipment such as a multichannel pipette and 96-well plates. Our protocol enables a single individual to construct all 256 possible consortia from a pool of eight species within an hour, and can be easily expanded to even higher-diversity libraries. Even without the aid of specialized equipment such as robotic liquid handlers, this procedure will make it possible to efficiently generate all communities with up to 10-12 species.

For laboratories equipped with robotic liquid handlers, the protocol can be easily expanded to accommodate larger libraries, with the main limitation being the amount of plasticware that will be used. The rationale underlying our assembly protocol allows for flexible implementation depending on the specific needs of each laboratory (e.g., it is straightfor-ward to generalize for 384-well plates). Our implementation also has the advantage of keeping pipetting error minimal and independent of consortium size, as we showed by creating all 256 color combinations from eight synthetic colorants. To facilitate automation and adaptation to the needs of other groups, we provide an accompanying R script to facilitate the generation of comprehensive step-by-step protocols for any number of species and pipetting volumes.

We have demonstrated the utility of this protocol in synthetic microbial ecology by generating all possible communities from a collection of eight P. aeruginosa strains and quantifying the strength of pairwise and higher-order functional interactions between them. Our experiment illustrates how the full factorial design of microbial consortia enables for the full characterization of ecological community-function landscapes, which can provide valuable insights into the emergence of community-level functions (such as, but not limited to, biomass productivity, as we studied here). The entire experiment, from spreading colonies to measuring community function, could be completed from beginning to end by a single experimentalist in less than one working week, and most of that time was taken by waiting for cell growth.

Although initially designed for building combinations of microbial species, our protocol can be readily applied for creating combinations of any soluble (or at least homogeneously distributed) compounds beyond microorganisms. For instance, it could be employed to generate all possible media compositions from a set of resources, all possible combinations of a set of antibiotics, toxins, or even bacteriophages, as well as mixed combinations thereof. We therefore anticipate that our methodology will find extensive applications and utility across diverse fields such as microbiology, ecology, evolution, and biotechnology.

Supporting information

eLife Assessment

This manuscript introduces a new low-cost and accessible method for assembling combinatorially complete microbial consortia using basic laboratory equipment, which is a valuable contribution to the field of microbial ecology and biotechnology. The evidence presented is convincing, demonstrating the method's effectiveness through empirical testing on both synthetic colorants and Pseudomonas aeruginosa strains.

Reviewer #1 (Public review):

Summary:

This work develops a simple, rapid, low-cost methodology for assembling combinatorially complete microbial consortia using basic laboratory equipment. The motivation behind this work is to make the study of microbial community interactions more accessible to laboratories that lack specialized equipment such as robotic liquid handlers or microfluidic devices. The method was tested on a library of Pseudomonas aeruginosa strains to demonstrate its practicality and effectiveness. It provided a means to explore the complex functional interactions within microbial communities and identify optimal consortia for specific functions, such as biomass production.

Strengths:

The primary strength of this manuscript lies in its accessibility and practicality. The method proposed by the authors allows any laboratory with standard equipment, such as multichannel pipettes and 96-well plates, to readily construct all possible combinations of microbial consortia from a given set of species. This greatly enhances access to full factorial designs, which were previously limited to labs with advanced technology.

Another strength of the manuscript is the measurement and analysis of the biomass of all possible combinations of 8 strains of P. aeruginosa. This analysis provides a concrete example of how the authors' new methodology can be used to identify the best-performing communities and map pairwise and higher-order functional interactions.

Notably, the authors do exceptionally well in providing a thorough description of the methodology, including detailed protocols and an R script for customizing the method to different experimental needs. This enhances the reproducibility and adaptability of the methodology, making it a valuable resource for researchers wishing to adopt this methodology.

Weaknesses:

While the methodology is robust and well-presented, there are some limitations that should be acknowledged more thoroughly. First, the method's scalability is an important factor. The authors indicate that it should be effective for up to 10-12 species, but there is no discussion of what sets this scale: time, amount of labor, consumables, the likelihood of error, sample volume, etc. Second, this methodology is tailored to construct communities where the abundance of each strain is identical in each combination. Therefore, combinations with a different number of strains also differ in the total initial amount of microbial cells. Second, variations in the initial proportions of the same set of strains cannot be readily explored. Third, the manuscript only discusses how to construct the combinations, and not how to assay them afterward (e.g. for community function, interspecific interactions, etc'). While details on how to achieve these goals are clearly outside the scope of this work, the use of biomass as an example function may obfuscate this caveat, which should be stated more explicitly.

Reviewer #2 (Public review):

Summary:

A simple and effective method for combinatorial assembly of microbes in synthetic communities of <12 species.

Strengths:

Overall this manuscript is a useful contribution. The efficiency of the method and clarity of the presentation is a strength. It is well-written and easy to follow. The figures are great, the pedagogical narrative is crisp. I can imagine the method being used in lots of other contexts too.

Weaknesses:

The authors could better clarify what HOIs mean. They could address challenges with assaying community function. However, neither of these "weaknesses" affects the primary goal of the paper which is methodological.

Reviewer #3 (Public review):

The authors developed a useful methodology for generating all combinations of multiple reagents using standard lab equipment. This methodology has clear uses for studying microbial ecology as they demonstrated. The methodology will likely be useful for other types of experiments that require exhaustive testing of all possible combinations of a given set of reagents (e.g., drug-drug antagonism and synergy).

The authors provided a useful R script that generates a detailed experimental protocol for building the desired combination from any number of reagents. The produced document is useful and has clear instructions. The output of the computer script will be strengthened if graphical output is also provided (similar to the one provided in Figure 1C).

The authors show that the error rate of the method doesn't go up with the number of combinations using dyes (Figure 2).

The authors demonstrate the value of their methodology for studying interactions within microbial consortia by assembling all possible combinations of eight strains of Pseudomonas aeruginosa. The value of their methodology for this application is well-founded. However, it is also unclear why specific experimental choices were made for this application. It is unclear why authors continue to show the absorbance measurements of strain assemblies over the entire wavelength spectrum and not just for ABS 600 nm (Figures 3 and 4). It is also unclear why the authors provided information on the "sum of the three spectra" as this reference line is meaningless and not a reasonable null model for estimating how well specific strain combinations will grow together.

Figure 5 illustrates the various analysis types that can be performed on the data collected from growing combinations of eight Pseudomonas aeruginosa strains. It is a very informative figure since it provides a "roadmap" on the various ways in which the dataset produced can be explored. The information in Figures 5 and S6 will likely be very useful for a wide audience.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

This work develops a simple, rapid, low-cost methodology for assembling combinatorially complete microbial consortia using basic laboratory equipment. The motivation behind this work is to make the study of microbial community interactions more accessible to laboratories that lack specialized equipment such as robotic liquid handlers or microfluidic devices. The method was tested on a library of Pseudomonas aeruginosa strains to demonstrate its practicality and effectiveness. It provided a means to explore the complex functional interactions within microbial communities and identify optimal consortia for specific functions, such as biomass production.

The primary strength of this manuscript lies in its accessibility and practicality. The method proposed by the authors allows any laboratory with standard equipment, such as multichannel pipettes and 96-well plates, to readily construct all possible combinations of microbial consortia from a given set of species. This greatly enhances access to full factorial designs, which were previously limited to labs with advanced technology.

Another strength of the manuscript is the measurement and analysis of the biomass of all possible combinations of 8 strains of P. aeruginosa. This analysis provides a concrete example of how the authors' new methodology can be used to identify the best-performing communities and map pairwise and higher-order functional interactions.

Notably, the authors do exceptionally well in providing a thorough description of the methodology, including detailed protocols and an R script for customizing the method to different experimental needs. This enhances the reproducibility and adaptability of the methodology, making it a valuable resource for researchers wishing to adopt this methodology.

We thank the reviewer for their thoughtful comments and positive assessment of our work. Below we detail the changes we have introduced in the manuscript to clarify issues raised by the reviewer.

While the methodology is robust and well-presented, there are some limitations that should be acknowledged more thoroughly. First, the method's scalability is an important factor. The authors indicate that it should be effective for up to 10-12 species, but there is no discussion of what sets this scale: time, amount of labor, consumables, the likelihood of error, sample volume, etc.

The 10-12 species estimation is based on our own experience implementing the protocol, and set primarily by time, labor, and consumables (as rightly pointed out by the reviewer) rather than conceptual limitations of the approach. We have added clarifications in the Discussion (lines 401-405) regarding these scalability-limiting factors.

Second, this methodology is tailored to construct communities where the abundance of each strain is identical in each combination. Therefore, combinations with a different number of strains also differ in the total initial amount of microbial cells. Second, variations in the initial proportions of the same set of strains cannot be readily explored.

Note that the “density homogenization” step is optional and it could be skipped entirely, which would result in a same species being present at variable densities across consortia: specifically, skipping this step would make the density of a species in a consortium inversely proportional to the number of species in that consortium. Further variations in initial abundance could be explored by treating a same strain at two (or more) starting abundances as distinct inputs of the protocol – though this would naturally increase the number of combinations to test.

We have included a paragraph in the Discussion (lines 416-423) describing how we can, in principle, extend our protocol to explore abundance effects.

Third, the manuscript only discusses how to construct the combinations, and not how to assay them afterward (e.g. for community function, interspecific interactions, etc.). While details on how to achieve these goals are clearly outside the scope of this work, the use of biomass as an example function may obfuscate this caveat, which should be stated more explicitly.

We agree that the manuscript focuses exclusively on the construction of microbial communities and does not address how these communities should be assayed afterward. This is an intentional scope decision. The proposed protocol is fully compatible with a wide range of functional, interaction-based, or omics-based assays. Absorbance is mentioned as an illustrative example of a possible readout, rather than as a recommended or exclusive parameter. We have revised the text to explicitly state that the assessment of community function or interspecific interactions lies outside the scope of this work and must be tailored to the specific biological question being addressed.

Reviewer #1 (Recommendations for the authors):

A few specific technical notes and notes about clarity:

(1) It may be worth being more explicit about how to produce replicates. For example, producing technical replicates by inoculating multiple times from the same set of combinations, while biological replicates require making the combinations multiple times.

We have updated the main text to clarify this point (line 780-781).

(2) Figure 2C: May be worth adding some context to these performance numbers. What are typical accuracies? What would they be in a liquid handler?

Assessing typical accuracies is nuanced since the error depends not only on the assembly steps, but also on potential intrinsic variation of the specific community function being tested and the method used to quantify it. One of the main reasons for including the experiment using colorant combinations was precisely to minimize these other sources of variation. In this experiment, we find that the error we quantify is consistent with cumulative pipetting variation (as a reference, a typical lab micropipette has an error of 0.5-1%). This is now explicitly mentioned in the manuscript.

(3) Figure 5A: I realize it is unlikely that strains go extinct in these experiments. But it is still worth clarifying that the number of strains is the number inoculated, rather than the one present at the time of measurement.

We updated the caption of Figure 5A as recommended by the reviewer.

(4) Figure 5B: I realize this is just for illustration purposes, but you should provide more information about the magnitude of the difference in performance of these combinations and the confidence in their ranking (or variability in performance across replicates).

Following this suggestion, we have added a paragraph where we report the variation across replicates for the highest-performing consortia (lines 318-323). Indeed, while variation across replicates is small, it is enough to produce an overlap between the confidence intervals of the function of some of the highest-performing consortia. This is now explicitly acknowledged in the manuscript.

(5) Figure 5C: I believe the bold black lines indicate the combinations shown in panel D, but that is not explicitly stated.

We have updated the caption of Figure 5C.

Reviewer #2 (Public review):

A simple and effective method for combinatorial assembly of microbes in synthetic communities of <12 species.

Overall, this manuscript is a useful contribution. The efficiency of the method and clarity of the presentation is a strength. It is well-written and easy to follow. The figures are great, the pedagogical narrative is crisp. I can imagine the method being used in lots of other contexts too.

The authors could better clarify what HOIs mean. They could address challenges with assaying community function. However, neither of these “weaknesses” affects the primary goal of the paper which is methodological.

We thank the reviewer for the positive assessment. With respect to HOIs, we recognize that defining and quantifying them is a non-trivial subject within the broader field of microbial ecology (see e.g. ref. 24 within the manuscript). Since our aim with this manuscript is methodological, as the reviewer notes, here we have done our best to avoid introducing new or ambiguous definitions. For this reason, we simply adopt a definition given in previous works (including refs. 10, 19, 24, 29, 37, and 38 in the manuscript), where the context-dependence of pairwise interaction terms is taken as a signature of HOIs. With respect to the challenges in assaying community function, please see our responses below.

Reviewer #2 (Recommendations for the authors):

Overall, this manuscript is a useful contribution, I appreciate the authors taking the time to write it up! I have a few relatively minor comments.

(1) It would be nice in the introduction to address why we might want the full factorial construction of communities in the first place. This is an especially relevant question in light of the authors' 2023 Nat E&E paper where they showed that the function of communities can often be learned even when only a fraction of all possible communities is measured. This is addressed in part in the paragraph on line 34, but I think it might be worth expanding a bit given the focus on the paper.

We sincerely appreciate the reviewer’s feedback. In fact, one of the reasons that make full factorial construction desirable is precisely to test theoretical and computational models of community function, including (but not only) the statistical models developed in our 2023 Nature E&E paper. In that work, we showed that low-order models can explain a substantial fraction of the variation in community function in previously-published datasets, but we also predict that the same models could fail under complex structures of microbial interactions (e.g., strong high-order interactions). The protocol we present here enables the empirical quantification of such interactions, making this prediction (and others) directly testable. We have included that clarification in the revised manuscript (lines 56-58).

(2) Around line 74, I think it is worth mentioning that even this elegant design will face insurmountable practical challenges (time, liquid handling operations, number of plates will explode) for full factorial design with 20, 30, 40 species or more. This is relevant for some very complex synthetic consortia that some microbiome groups are constructing (e.g. hCom2 from Huang/Fishbach groups) https://www.sciencedirect.com/science/article/pii/S0092867422009904.

We agree with the reviewer that full factorial designs become impractical for very large species pools. These limits are now more clearly mentioned in the revised manuscript. We refer the reviewer to our response to comment #1 by Reviewer 1 for further details.

(3) The binary construction is a really nice clean way to explain the protocol. Appreciate the pedagogy!

We thank the reviewer for the appreciation.

(4) In the experiment with pseudomonas strains the consortia are grown in LB. This medium will support growth to relatively high OD (>1). At these densities, the change in OD with density is almost certainly not linear with cell density, and this nonlinearity likely depends on strain identity. In this case, the assumption of additivity may not hold. As a result, some of the observed "interactions" may simply be non-linearity in the assay and not the abundance of bacteria in the communities. Of course, this does not affect the assembly protocol in any way, but it does complicate the interpretation of interactions via this assay. I think this is worth pointing out since other researchers may have to think carefully about the assay they use when constructing these synthetic consortia. I think in this methods paper it is important to emphasize this so other researchers do not mistakenly identify interactions due to issues with the assay.

We thank the reviewer for pointing out this important aspect. In our experiment, we use Abs600 simply as an example of a measurable community-level function. The reviewer is absolutely correct in that mapping absorbance to biomass is nuanced at large OD values, where this relationship becomes non-linear. While this is not an issue from the perspective of the protocol itself, it is indeed an important consideration for users who may want to obtain reliable quantifications of biomass. We have updated the manuscript to explicitly mention this potential issue (lines 307-313). We have also emphasized the fact that our focus on Abs600 is strictly for illustrative purposes, and we have removed all instances where a direct mapping from Abs600 to biomass was implied in the text.

(5) Subtle point regarding HOIs. HOI (or pairwise) statistical interactions need not quantitatively be the same as interactions in a lotka volterra sense. I realize the authors do not explicitly use the term "interaction" in an gLV model formalism but this is how the majority of readers will interpret this term. I believe it is a research question as to how pairwise gLV interactions manifest themselves in terms of functional interactions. For example, a purely pairwise LV model could easily have HOI "functional interactions" if the function is total abundance since abundances depend nonlinearly on LV interactions. I think this part of the manuscript could be confusing to readers for this reason. I think the term "functional interaction" really helps with this issue, but just asking the authors to make sure this is clear.

I say this because ref: 37 is focused on HOIs in an LV sense. Here, as the authors are aware, they are computing statistical "interactions" in the sense of epistasis. Given that they are computing this epistasis averaged across all community compositions a more appropriate citation might be [https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004771] where the same quantity is computed in a protein context.

We thank the reviewer for pointing out this important issue. Indeed, we use the term “interaction” in a statistical sense (as the deviation of the observed community function from a null, additive expectation) rather than in a Lotka-Volterra sense. We agree that the reference suggested by the reviewer is more appropriate in this context. We have updated the reference list accordingly.

(6) Figure 5G - a little hard to see. Any way to show this data more clearly? It looks like all interactions have a mean of 0 because of the way the data are presented.

The reviewer is indeed correct in that, as defined, the interactions that we quantify are back ground dependent, and their average across backgrounds lies near zero for all species. More than an issue with the representation, we think that this is an important empirical observation: it indicates that a same species pair may interact positively or negatively depending on its ecological context. We believe that the current representation is most appropriate for making this clear, but we would be open to discussing alternatives if the reviewer had a specific suggestion in mind.

Reviewer #3 (Public review):

The authors developed a useful methodology for generating all combinations of multiple reagents using standard lab equipment. This methodology has clear uses for studying microbial ecology as they demonstrated. The methodology will likely be useful for other types of experiments that require exhaustive testing of all possible combinations of a given set of reagents (e.g., drug-drug antagonism and synergy).

The authors provided a useful R script that generates a detailed experimental protocol for building the desired combination from any number of reagents. The produced document is useful and has clear instructions. The output of the computer script will be strengthened if graphical output is also provided (similar to the one provided in Figure 1C).

The authors show that the error rate of the method doesn't go up with the number of combinations using dyes (Figure 2).

The authors demonstrate the value of their methodology for studying interactions within microbial consortia by assembling all possible combinations of eight strains of Pseudomonas aeruginosa. The value of their methodology for this application is well-founded. However, it is also unclear why specific experimental choices were made for this application. It is unclear why authors continue to show the absorbance measurements of strain assemblies over the entire wavelength spectrum and not just for ABS 600 nm (Figures 3 and 4). It is also unclear why the authors provided information on the "sum of the three spectra" as this reference line is meaningless and not a reasonable null model for estimating how well specific strain combinations will grow together.

Figure 5 illustrates the various analysis types that can be performed on the data collected from growing combinations of eight Pseudomonas aeruginosa strains. It is a very informative figure since it provides a "roadmap" on the various ways in which the dataset produced can be explored. The information in Figures 5 and S6 will likely be very useful for a wide audience.

Reviewer #3 (Recommendations for the authors):

(1) Congratulations. I think the manuscript lays out a simple and very elegant methodology that will be useful for many. While I think the method is overall well explained and rationalized, the paper can greatly benefit from further expansion of Figure 5 at the expense of Figures 3 and 4.

We thank the reviewer for their thoughtful assessment of our work. We have considered the recommendations and discuss the following points in response.

(2) Unless I am missing something, there is no reason to present data collected across the entire wavelength spectrum for microbial assemblies (Figures 3 and 4). Moreover, using the same color palette for bacterial strains (Figure 3A) and colorants (Figure 2) is highly confusing. I suggest considering using only the 600 nm wavelength for any data collected from microbial assemblies and using a very different color palette for bacteria and colorants to avoid misinterpretation of the data.

We thank the reviewer for this suggestion. Our goal with Figures 3-4 was to illustrate the convenience of the protocol and the ease with which many measurements can be performed in parallel once the combinatorial assembly has been completed. While we focus on Abs600 for all subsequent analyses, we chose to display the full spectra in Figs. 3-4 in hopes that future studies can make use of our rich dataset to interrogate questions on microbial interactions, with the option to focus on other wavelengths (which can effectively be treated as different community-level functions in their own right; for instance, we have previously used Abs405 as a proxy for siderophore concentration). We think there is value in Figs. 3-4 in their current form to make this clear to readers.

(3) Unlike dye absorbance, bacterial carrying capacity has an upper limit, so summing individual population absorbance as a reference line seems unjustified. If the summation of absorbance is meant to provide a "null model" for expected growth, a more suitable model should be considered (e.g., max spectra or a weighted sum of the spectra from individual members).

We agree with the reviewer that our null model is not biologically constrained, and we did not intend to imply that the additive expectation was derived from biological principles. Instead, this additive expectation should be interpreted as a simple statistical baseline with minimal assumptions. The use of an additive baseline for quantifying microbial interactions has been addressed in the literature (see, e.g., references 10, 19, 24, 29, 37, and 38), and so here we chose to conform to this convention to avoid introducing new, non-standard quantifications of pairwise and higher-order interactions. We have revised the text to make this more explicit.

(4) The R script is a valuable tool. I think that a valuable improvement will be to also generate visual representations as part of the script’s output such as the colored plates in Figure 1C that are specific to the generated protocol.

We have updated the script so that it now also outputs a table specifying the location of each consortium within the plates. We chose to make this a text, rather than a graphics output, to ensure cross-device compatibility.

(5) The discussion rightly acknowledges the potential to extend the protocol to larger libraries using liquid handlers. To facilitate this implementation, it might be beneficial to modify the script output so that the ‘volume’, ‘plate’, and ‘column’ values are tab- or comma-delimited.

We thank the reviewer for the suggestion. We have modified the output so that it is now tab-delimited.

(6) Figures 3 and 4 do not provide a lot of insight. I would suggest combining them into a single figure and using only absorbance values at 600 nm. It would also be interesting to add a histogram of these absorbance values and possibly show histograms for subgroups (e.g. all assemblies with more than 3 strains vs all assemblies with 3 or fewer strains).

With respect to Figs. 3 and 4, we refer the reviewer to our response to comment #2. With respect to the histogram/subgroups plot, we understand that this would be a slightly modified version of the current Fig. 5A, where we show means and standard deviations across all subgroups of 1 to 8 species, and so we find it unclear what this figure would add.

(7) With the recommendations of removing or reworking Figures 3 and 4, and the fact that Figure 5 is data-rich (and extremely useful), it would be beneficial to split Figure 5 and include the data shown in Figure S6 in the main figure. The analysis in Figure 6S is valuable and it might be beneficial to elevate this analysis to a primary figure and provide a detailed explanation of its rationale and methods in the main text.

We appreciate this suggestion. In our view, we find that both the text and the figures benefit from a heavy focus on the assembly protocol, as this is the main contribution of this work. While we do think it is valuable to highlight the type and amount of data that can be collected with a full factorial assembly, as well as the types of analyses that can be performed with this data, we are afraid that allocating more space to these analyses may distract readers from the methodology itself. We have therefore chosen to keep the original structure for Figs. 5 and S6.

Version 2

Abstract

Constructing combinatorially complete species assemblages is often necessary to dissect the complexity of microbial interactions and to find optimal microbial consortia. At the moment, this is accomplished through either painstaking, labor intensive liquid handling procedures, or through the use of state-of-the-art microfluidic devices. Here we present a simple, rapid, low-cost, and highly accessible liquid handling methodology for assembling all possible combinations of a library of microbial strains, which can be implemented with basic laboratory equipment. To demonstrate the usefulness of this methodology, we construct a combinatorially complete set of consortia from a library of eight Pseudomonas aeruginosa strains, and empirically measure the community-function landscape of biomass productivity, identify the highest yield community, and dissect the interactions that lead to its optimal function. This easy to implement, inexpensive methodology will make the assembly of combinatorially complete microbial consortia easily accessible for all laboratories.

Introduction

Microbial consortia are playing a growingly important role in biotechnology. They possess substantial potential advantages over monocultures, including larger metabolic capabilities, division of labor, and a potentially higher ecological and evolutionary stability. 1 Synthetic microbial communities are being engineered for goals as diverse as degrading pollutants, 2 producing high-value molecules such as biofuels, 3,4 vitamins, 5,6 and flavonoids, 7,8 and preventing the invasion of pathogens. 9

To fully realize the promise of microbial consortia in biotechnology we must develop tools for their optimization. Given a library of candidate strains, which consortia should we form if we wish to maximize a function of interest? 10 One way to address this question is purely empirical, following a full factorial design in which we form every possible assemblage and simply select whichever one is best. 11–15 Even for candidate libraries of small size (less than 10 species) the combinatorial nature of this strategy makes the assembly of all possible communities painstaking and tedious, limiting the scale and replicability of full factorial design. The number of unique liquid handling events required to form all possible combinations of m species scales as m 2m−1, as each of the m species needs to be added to each of the 2m−1 consortia where it is present. Manually making all combinations is therefore draining, slow, and prone to human error. Because of the long time required to complete such a task, the risk of contamination is high, and replication is constrained. Perhaps for this reason, studies reporting full factorial construction of microbial consortia are very rare, 11,13,14,16 and the field has largely relied on fractional factorial design, where only a subset of representative species combinations are constructed. 10,17–19

The ability to form full factorial combinations of a library of species would have benefits beyond biotechnology. In ecology, forming every possible species assemblage is required if we wish to investigate the complex, combinatorial nature of microbial interactions 16,20–25 and the role that high-order interactions play in the function and dynamics of microbial communities. 19,24,26–31 It would also allow for the testing and validation of theoretical and computational models of community function. 10,24,25,31 In sum, the quantitative understanding of high-order microbial interactions, the empirical mapping of community assembly networks, the quantitative exploration of community-function landscapes, and the development of predictive models of microbial community assembly and function all require us to be able to reliably assemble communities of growing complexity. 15,18,31

Robotic liquid handlers can facilitate the task of assembling a full combinatorial set of microbial communities. 11 While this alleviates some of the challenges associated to combinatorial liquid handling, particularly those associated with human error, others remain: the execution of hundreds or thousands of pipetting events is still inevitably slow, and as a result additional equipment (i.e., a HEPA filter) is required to avoid contamination. Robotic liquid handlers are also expensive, technically sophisticated equipment and they are not routinely available. As an alternative to deal with the combinatorial complexity of community assembly, Blainey et al. have introduced a droplet-based microfluidic system (kChip) that is capable of forming hundreds of thousands of species assemblages, 27,32 facilitating their full factorial construction. Notwithstanding their power and unparalleled high-throughput, microdroplet-based approaches such as kChip and others 33 are still state-of-the-art experimental methods, requiring specialized equipment and training that are not yet available to the vast majority of research groups worldwide.

In this paper, we describe a simple, rapid, inexpensive, and highly accessible liquid handling methodology for the full factorial design of microbial consortia and inocula, which requires only basic laboratory equipment and which can be easily implemented by the vast majority of research laboratories. Using a standard multichannel pipette, a single user can manually assemble all possible combinations of up to 10 species in less than one hour, a timescale that is shorter than the replication time of most bacteria in minimal media. Higher-dimensional designs are possible too, with the only limitation being the plasticware that is required. In addition to providing detailed protocols for a specific set of conditions (e.g., volume of the culture, density of the bacteria, etc.) we provide an R script to help the user tailor the protocol to their needs (Supplementary file 1). To illustrate the usefulness of this approach, we apply it to empirically characterize the full, combinatorially complete community-function landscape of a model microbial consortium consisting of eight Pseudomonas aeruginosa strains. We characterize the full absorption spectrum of all consortia and, as a model function, we determine their total biomass. We show that the implementation of our protocol, enabling the full factorial assembly of all strain combinations, allows us to quantitatively determine the relationship between community diversity and function, identify optimal strain combinations, and characterize all pairwise and higher-order interactions among all members of the consortia. We expect this methodology will rapidly expand the number of factorially constructed microbial consortia in the literature.

Materials and methods

In what follows, we describe the reasoning behind our protocol in full detail. Readers interested in the practical implementation of this logic are referred to Appendix 1, where we provide a step-by-step guide for the assembly of all consortia from a particular example, consisting of a library of eight microbial isolates. In addition, we provide an R script (Supplementary file 1) which can be used to generate similar step-by-step guides for different numbers of species and pipetting volumes.

Logic of the assembly protocol

The mathematical basis of our method lies in identifying each microbial consortia by a unique binary number. For a set of m species, we can represent any consortium (which we generically call c) as (1) c=xmxm−1xm−2⋯x2x1 where xk = 0,1 represents the absence (0) or presence (1) of species k in the consortium (Figure 1A). For example, for m = 6 species, the consortium where only species 1, 2, and 3 are present would be represented as 000111.

Logic of the assembly protocol.

(A) We consider a set of m species (in this illustration, m = 8), and we identify each of the 2m consortia that can be formed with them by the binary number c = xmxm−1 ⋯ x2x1 (with xk = 1,0 representing the presence/absence of species k, respectively; see equation 1). (B) Our protocol is based on the idea that one could assemble each of the 23 = 8 combinations of species 1 to 3 in the first column of a 96-well plate. Iteratively duplicating these consortia and adding species 4 to m to them would eventually result in all 2m combinations of species, as described in the main text. (C) Each of the m species would be present in the indicated wells (see equation 2). (D) Positioning of the 2m consortia within the 96-well plates used in the protocol.

Binary numbers are the natural way to denote combinations in mathematics, and indeed the correspondence between the two sets is often used in combinatorics. For instance, there are (mn) combinations of m species. The same number counts how many binary strings of length m contain n ones (equivalently, n zeros).

Now, the most important aspect of this notation is that merging two disjoint consortia becomes a simple addition: if we combine consortium 110000 with consortium 000011, the resulting consortium is represented by the binary number 110011, which is the sum of the binary representations of the initial consortia: 110000 + 000011 = 110011. Note, however, that this operation only works for disjoint consortia. If we add two consortia sharing one or more species, binary addition is no longer a good representation: combining consortium 100001 with consortium 000001 should result in the first consortium again, 100001 (we are for now ignoring potential differences in species abundances/densities, but will address this issue later in the text). Yet, the addition of the two binary representations leads to 100010, which does not represent the appropriate consortium (remember that in binary addition 01 + 01 = 10). As we will see in what follows, our protocol minimizes the number of liquid handling events by making extensive use of this addition property, but only for disjoint consortia.

The second property that we will use to our advantage is that 96-well plates have 8 rows, which is a power of 2 (23 = 8). This allows us to form all combinations from a 3-species set (say, species 1, 2, and 3) in one column of the plate. It is convenient to arrange these 8 consortia following the order of their binary representation: the empty consortium (000) in the first well, followed by 001, 010, 011, 100, 101, 110, and 111; corresponding to the decimal numbers from 0 to 7 in increasing order.

We can now duplicate these eight consortia, pipetting them into the second column of the 96-well plate, and then pipette species 4 (1000) into all wells of this second column using a multichannel pipette (Figure 1B). This operation is equivalent to the binary addition of consortium 1000 (species 4 alone) with each of the starting eight consortia, and results in all 24 = 16 consortia which can be assembled with species 1 to 4: in the first column, as explained before, we have 0000, 0001, 0010, 0011, 0100, 0101, 0110, and 0111 (decimal numbers 0 to 7); while in the second column we now have 1000, 1001, 1010, 1011, 1100, 1101, 1110, and 1111 (decimal numbers 8 to 15, Figure 1B). We can next duplicate each of these 16 consortia, pipetting the first and second columns of the plate into the third and fourth columns, respectively. If we then add species 5 (10000) to each of the duplicated consortia, we generate all 25 = 32 combinations of species 1 to 5 (Figure 1B).

In what follows, we will describe how this logic can be implemented most efficiently in practice, but hopefully the reader can already intuit how to generalize this algorithm for consortia with more species. Note that for more than 6 species, multiple 96-well plates will be required: for instance, with m = 7 one could form 27 = 128 potential consortia, more than what a single 96-well plate can fit. Although we will not make use of them here, the algorithm is also generalizable for 384-well plates, as these have 16 rows. In this case, we would use the first column to assemble all consortia made up by species 1 to 4, and then proceed as described above for all subsequent species.

Plate arrangement

Following the logic we just discussed, it is straightforward to notice that species 4 will be present in all wells of columns 2, 4, 6, 8… of the 96-well plate. In turn, species 5 will be present in alternating pairs of columns: 3, 4, 7, 8, 11, 12… Species 6 will be present in alternating sets of 4 columns: 5, 6, 7, 8, 13, 14, 15, 16… (Figure 1C). Note that 96-well plates only have 12 columns: what we here call columns 13 to 24 would in practice be columns 1 to 12 of a second 96-well plate, columns 25 to 36 would correspond to a third plate, and so on. By a similar reasoning, species 1 to 3 will be present in alternating sets of 1, 2, and 4 rows (instead of columns) respectively (Figure 1C). In summary, we have that species k will be present in: (2) Species k present in:{alternating sets of 2k−1 rows if k=1,2,3alternating sets of 2k−4 columns if k=4,5,⋯,m

Conveniently, this arrangement makes it so consortia are located in order from top to bottom and from left to right within the plate(s): the first column contains the consortia corresponding to the binary representations of decimal integers 0 (00000) to 7 (00111), the second column corresponds to decimal integers 8 (01000) to 15 (01111), the third column to decimal integers 16 (10000) to 23 (10111), and so on.

Homogenizing species’ densities across consortia

As we just saw, it would be possible to form all combinations of m species simply by pipetting each one of them row-wise (for species 1 to 3) or column-wise (for species 4 to m) in their corresponding positions (equation 2) with the aid of multichannel pipettes. There is, however, one important caveat: if we pipette equal volumes of each species’ monocultures, we will end up with variable volumes across wells. As an example, if we pipette a volume v0 of each monoculture, we will have a volume of 2v0 in well D01 of the 96-well plate (where consortium 011 is located), but a volume of 3v0 in well H01 (where consortium 111 is located). Species 1 is present in both of these example consortia at the same total population size, but at different population densities differing by a factor of 2/3.

Often, we may want the density of each species to be consistent across consortia. To achieve this, we can pipette additional liquid (sterile water, growth medium, saline buffer…) into each well in order to compensate for the differences in volume. Note that, prior to this density homogenization step, the maximum volume in a well would be mv0, corresponding to the consortium with all m species. This naturally also corresponds to the consortium where each of the individual species is present at the lowest population density. In every other consortium (say, consortium c), the volume of buffer (which we denote vB(c)) that we would need to add in order to reach this minimum population density is given by (3) vB(c)=v0(m−H(c)) where H(c) represents the Hamming weight of the binary representation of consortium c. The Hamming weight of a binary number is the number of ones in its representation, for instance, H(0000) = 0, H(0010) = 1, H(1111) = 4, etc. Thus, H(c) simply counts the number of species present in consortium c.

In practice, we do not need to manually pipette the buffer well-by-well. The convenient arrangement of the plate allows us to streamline this process. In what follows, we use i and j as the row and column indexes of the plate (from top to bottom and from left to right, respectively). For instance, well C02 (containing consortium 1010) would correspond to i = 3 and j = 2. As a reminder, we are considering that j = 13 to 24 correspond to columns 1-12 of a second 96-well plate, j = 25 to 36 correspond to a third plate, etc.

In order to streamline the pipetting of buffer, it would be convenient to find an expression for H(c) in terms of the positional indexes i and j. Now, due to the particular arrangement of consortia in the plate, we can divide the binary representation of a consortium c into two. First, the right-most three bits of c are determined by the row in which it is located. For instance, all consortia in row 7 will always have a binary representation ending in 110, independently of the column in which they appear. In general, the three rightmost digits of c will be equal to B(i − 1), the binary representation of i − 1 with i being the index of the row where consortium c is located. Similarly, the leftmost m − 3 bits of c will depend only on the column, in the following way: take a column j and ignore the three rightmost digits; then the binary representation of all consortia in that column will be the same, and equal to B(j − 1). For instance, with m = 8, the binary representation of all consortia in column 4 is 00011x3x2x1 — the last three bits may be 0 or 1 depending on the specific well, but the first five bits are common to all wells in the column. Therefore, in order to calculate the Hamming weight of a given consortium, we just need to sum the Hamming weight of the row and the Hamming weight of the column: (4) H(c)=H(B(i−1))+H(B(j−1))

We can use equation 4 to rewrite equation 3 in terms of the positional indexes i and j. After some algebra, we arrive at (5) vB(i,j)=v0[3−H(B(i−1))]+v0[m−3−H(B(j−1))]

While terms in this equation could be arranged in a more compact form, equation 5 as it is allows us to implement the buffer pipetting in a very efficient way: We can start by pipetting liquid buffer column-wise (a volume v0[m − 3 − H(B(j − 1))] into every well of column j). We can then pipette the buffer row-wise, pipetting a volume v0[3 − H(B(i − 1))] into every well of row i. Furthermore, the row-wise buffer pipetting can be bypassed if we homogenize the densities of the starting consortia formed with species 1 to 3, as we explain in detail in Appendix 1.

Results

Proof of concept: construction of colorant combinations

As a first demonstration of the method, and to establish its feasibility and accuracy, we used commercial food colorants and temperas to build all 256 combinations made up from eight colors. We diluted these colorants in water such that all of them had comparable maximum absorbances in the range from 380 nm to 780 nm. Each color exhibits a different absorbance spectrum in this range (Figure 2A).

Full factorial construction of synthetic colorant combinations.

(A) We consider a set of eight synthetic colorants, each of which exhibits a different absorbance spectrum in the range 380-780 nm. (B) As an example, we show the empirically measured absorbance spectrum (solid black line) for the combination of colorants 1, 3, and 8 (10000101). The dashed black line represents the additive expectation, i.e., the sum of the spectra of the three constituent colorants (colored dashed lines). (C) Relative deviation between the empirical spectrum and the additive expectation (see equation 6) for the example combination 10000101 (left) and for all mixtures of 2 or more colorants (right). Relative errors were only computed when the true absorbance was above 0.1 A.U.

Using synthetic colorants to test our protocol has various advantages. First, color differences are clearly visible to the naked eye, making it easy to keep track of the location of each colorant. Second, we do not expect these colorants to interact with one another, or to do so very weakly. This means that the absorbance spectrum of any combination of colorants may be reasonably expected to match the sum of the individual spectra of each colorant. Therefore, the magnitude of the deviations between an empirical spectrum and this additive expectation (Figure 2B) can be used to estimate an upper bound for the pipetting error of the protocol. We can obtain an estimation of the accuracy of our protocol by examining these deviations.

We followed the protocol described above, setting m = 8 and v0 = 25 μL (a detailed step-by-step protocol for these parameters is given in Appendix 1), to assemble all 256 color combinations. The absorbance spectrum of each combination was quantified using a MultiSkan SkyHigh plate reader (Thermo Scientific). The implementation of the protocol required three 96-well plates, 8 Falcon tubes, an 8-channel pipette and a 12-channel pipette. Completing the whole protocol took a single experimentalist less than an hour. Figure 2-figure supplement 1 shows the three 96-well plates at six different steps of the protocol.

Figure 2B shows the absorbance spectrum for an example colorant mixture (10000101), comparing it to the additive expectation that its three constituent colorants (1, 3, and 8) do not interact. We estimated the pipetting error of the protocol as the relative deviation (denoted δ) between the empirical spectra and the additive expectations at all wavelengths (Figure 2B), that is: δ(c,λ)=|Abλ(c)−Absλ(add)(c)|Absλ(c) where we have denoted Absλ(c) the absorbance at a wavelength λ (between 380 and 780 nm) of the colorant mixture c, and Absλ(add ) (c) the additive expectation (i.e., the sum of the absorbances of each individual constituent colorant).

In Figure 2C we show the distribution of relative deviations for the example mixture 10000101 at all wavelengths (left), as well as the distribution corresponding to all mixtures at all wavelengths (right). Figure 2-figure supplement 2 shows the absolute deviations for all mixtures at all wavelengths (median absolute deviation ∼ 0.025 A.U.). We found that relative deviations were generally small (Figure 2C: mean ∼ 5.8%, median ∼ 4.9%). As a reference, we note that typical lab micropipettes have an error of 0.5-1%, and so these numbers are compatible with cumulative pipetting variation. We also found that the magnitude of these deviations did not increase with the number of colorants in a mixture (Figure 2-figure supplement 3), suggesting that the precision of our protocol is not compromised by the size of the set from which mixtures are assembled. Deviations were also roughly similar in magnitude across all wavelengths (Figure 2-figure supplement 4).

Full factorial design of synthetic bacterial consortia

To illustrate the potential applications of our protocol for finding optimal microbial consortia and quantitatively studying the complexity of microbial interactions, we adopted a collection of eight Pseudomonas aeruginosa strains obtained from a previous experiment 34–36 (see Appendix 2 and Supplementary file 2 for details). Monocultures of each of these strains grown in liquid LB medium exhibit differences in their absorbance spectra (Figure 3A). Unlike the synthetic colorants, bacterial strains can be expected to interact with one another when co-cultured through a variety of potential mechanisms, including competition for limiting resources or facilitation which may lead to changes in the population sizes of different species. 15,19

Co-culturing combinations of bacterial isolates.

(A) We consider eight P. aeruginosa strains obtained from a previous experiment. 34–36 Here we represent the absorbance spectra of the eight monocultures of these strains. Solid lines and shaded areas represent means and standard deviations across three independent biological replicates. (B) We followed our protocol to assemble different combinations of strains (in this example, the consortium formed by strains 2, 3, and 8, represented as 10000110). Each consortium was diluted into fresh medium by a factor of 1:400 and allowed to re-grow for 40 h, after which the absorbance spectrum of the culture was measured. Solid black line: empirical spectrum of consortium 10000110 (mean of three independent biological replicates, shaded region represents the standard deviation across them). Dashed black line: additive expectation if the three strains did not interact. Dashed colored lines: monoculture spectra of the three constituent strains.

We first cultured each strain separately by resuspending a single colony into 15 mL of LB medium in 50 mL conical Falcon tubes. Strains were allowed to grow to carrying capacity for 24 h at 37 °C with no agitation. Then, monocultures were fully homogenized and used to assemble all 256 strain combinations following the same protocol described before (detailed steps can be found in Appendix 1). Phosphate buffer saline (1×PBS) was used to adjust volumes/densities as indicated in the protocol. After completing our protocol (which took a single experimentalist less than one hour), each consortium was inoculated into fresh LB medium (0.5 μL of inoculum into 200 μL of medium for a dilution factor of 1:400; note that our protocol design makes it so this is equivalent to inoculating 0.5 μL/8 = 0.0625 μL of each strain’s monoculture into the fresh medium) in 96-well plates. Samples were incubated still at 37 °C for 40 h, after which the absorbance spectrum of each sample was quantified in the same plate reader we used before.

Figure 3B shows the absorbance spectrum for one example consortium after the 40 h incubation period. Consistent with our reasoning, this spectrum deviates strongly from the additive expectation (sum of each monoculture spectrum), indicating a prominent effect of interactions between strains. Figure 4 shows the spectra for all consortia with two or more members, where deviations from the additive expectation are similarly large.

Full factorial design of synthetic microbial consortia.

We represent the absorbance spectra of all consortia (of 2 or more strains) that can be formed with our library of eight P. aeruginosa isolates. As in Figure 3B, solid black lines and shaded areas represent the mean and standard deviation of the co-culture spectrum across three biological replicates; the dashed colored lines represent the individual monoculture spectra; and dashed black lines represent the sum of these individual spectra.

Because we have constructed all possible assemblages, we can systematically investigate the interactions between species and how they may result in different community-level properties and functions. As an illustration, we focus on the absorbance of the liquid cultures at 600 nm (Abs600). While this is a commonly used proxy for biomass in microbial cultures, here we use it simply as an example of a quantitative community-level function in itself. Note that the relationship between absorbance and biomass is only linear at low values of Abs600; readers interested in reliably quantifying biomass should keep this limitation in mind when designing experiments. We found a non-monotonic relationship between Abs600 and the number of inoculated strains, peaking at 3 strains (Figure 5A). Our protocol allows us not only to characterize this type of broad-scale relationships between diversity and function, but also to identify specific high-performing consortia (Figure 5B). In our experiment, the consortium with the highest function contains only three strains (consortium 00001101, containing strains 1, 3, and 4). Note that the ranking depicted in Figure 5B serves mere illustrative purposes, as in our data many consortia exhibited similarly high biomass values. As a reference, consortium 00001101 had Abs600 = 1.294 ± 0.048 A.U. (mean and standard deviation across n = 3 biological replicates; see Appendix 2), consortium 00000101 had Abs600 = 1.291 ± 0.066 A.U., and consortium 01100101 had Abs600 = 1.289 ± 0.05 A.U. In Figure 5C we represent the full map between the composition and function of all 256 consortia in our experiment (i.e., the community structure-function landscape 15,16,19,22).

Full factorial design of microbial consortia enables the dissection of microbial interactions.

We focus on the absorbance at 600 nm as a proxy for the biomass of a consortium. (A) Consortium function (biomass quantified as Abs600) against the number of strains inoculated into the consortium. Dots and error bars represent the average and standard deviation across all consortia with a same number of strains. (B) Top three consortia exhibiting the highest biomass across all 28 combinations. (C) We represent the function (Abs600; values shown are averages across three biological replicates) for the consortia assembled using our protocol. Nodes correspond to the different consortia; edges connect consortia that differ in the presence of a single particular strain. Bold lines correspond to the detail highlighted in panel D. (D) Detail showing the effect of including strain 1, strain 2, and both, in a background consortium (in this example, the background corresponds to consortium 11001000). Red: the functional effect of strain 1 is quantified as the difference in function between the background consortium and the consortium resulting from the inclusion of that strain. Green: the functional interaction between strains 1 and 2 in this particular background is quantified as the difference between the additive expectation (dashed line) and the empirical value for the function of the consortium containing both strains, as explained in the main text. (E) Detail of the functional landscape corresponding to the three strains that are part of the overall highest-biomass consortium (00001101). Inset: functional interactions within this 3-strain consortium. Order 1 corresponds to the monoculture functions. Order 2 corresponds to pairwise species-by-species interactions as defined in panel D of this figure. Order 3 corresponds to the third-order interaction between these strains, defined as the deviation between the empirical function of the trio with respect to the null expectation from the additive and pairwise interactions. 19,29,30 (F) Distributions of functional effects of the eight strains used in our experiments across all of the 128 backgrounds where each may be included. (G) Distributions of functional interactions between all pairs of strains across all backgrounds where both may be included. (H) Fraction of functional variance due to interactions of different orders (order 1 corresponds to the additive effects of each strain). The fractions were computed as explained in Appendix 4. 31 Dashed line is the cumulative sum. (I) We have recently shown that the functional effect of a strain is often predictable from the function of the background consortium through simple, linear regressions. 10 For the eight P. aeruginosa strains in our experiment, linear regressions do an excellent job at linking these functional effects with the function of the background consortia.

From this type of data, one can readily quantify the sign and magnitude of strain interactions with respect to the community-level function of interest, in our case Abs600. In Figure 5D we show that the inclusion of strain 1 in a given consortium (which we refer to as an ecological background) results in an increase in function, and the same is true for the inclusion of strain 2. We call the difference in function between two consortia with and without a given focal strain as the functional effect of that strain in that ecological background. 10 If two strains did not interact, we may expect that including both of them in a same ecological background would result in a change in function equal to the sum of their two individual functional effects (see Figure 5D dashed line represents this additive expectation). If the empirical function deviates from the additive expectation, the two strains engage in a functional interaction (denoted ϵ in Figure 5D) in that particular ecological background (see Appendix 3 for details). 10,19 Under this definition, the term interaction should be interpreted in a statistical, rather than biological sense, where the null additive expectation serves as a mere minimal-assumptions baseline that is not intended to capture any particular biological mechanism (e.g., competition for nutrients).

Following this definition, one could dissect the interaction structure of all eight P. aeruginosa strains in our system. As an illustration, in Figure 5E we show the community-function landscape corresponding to all consortia that could be formed solely with strains 1, 3, and 4, i.e., the trio that maximizes Abs600 (see Figure 5B-C indicating that the maximum Abs600 is achieved by consortium 00001101). Although pairwise interactions are negative within this consortium, the three strains engage in a positive third-order functional interaction (see inset in Figure 5E) which rescues the function of the trio. This observation highlights the potential of full factorial design to interrogate and generate hypotheses with respect to the mechanisms that govern the emergence of community-level functions, particularly (but not only) in high-performing consortia. Beyond this particular consortium, in Figure 5F we represent the distributions of functional effects of all eight strains in our library across all of the 128 ecological backgrounds where they may be included. We found that the functional effects of all eight P. aeruginosa strains exhibit large variation in sign and magnitude, indicating that the contribution of a strain to the total community function largely depends on the presence or absence of additional community members in the ecological background. This indicates a prominent role of inter-strain interactions in this system.

In Figure 5G we represent the magnitude of the functional interactions between all pairs of strains i and j (denoted ϵij) across all the ecological backgrounds where the two may be included. We found that these pairwise interactions are highly variable across backgrounds. Pairwise interactions being sensitive to the presence or absence of additional community members can be interpreted as a signature of higher-order ecological interactions (HOIs). 19,37,38 By this definition, HOIs do exist in our particular experimental system (as evidenced by the variation in the pairwise interaction terms ϵij shown in Figure 5G); however, most of the functional variance can be explained by low-order interactions (Figure 5H the fraction of functional variance attributed to pairwise and higher-order interactions was computed as described elsewhere; 31 see Appendix 4 and Figure 5-figure supplement 1 for details). This has also been shown to be the case for other datasets of microbial community function. 31 We have recently shown that the aggregate effect of pairwise and higher-order microbial interactions often results in emergent statistical patterns, which make it so the functional effect of a strain may be predictable from the function of the ecological background where that strain is included. 10 The relationships between the functional effect of a strain and the function of the background community can often be captured through simple linear models, which mirror the patterns of global epistasis commonly reported in genetics. This phenomenon is also observed in the particular set of P. aeruginosa strains that we analyzed here (Figure 5I R2 averages 0.54), each of which exhibits a different relationship between its functional effect and the background function. Adopting the same analytical approach we used in previous work, 10 we find that such strong global epistasis patterns deviate from the null expectation of a random community-function landscape (Figure 5-figure supplement 2).

Together, these analyses illustrate how a full factorial construction of microbial consortia may allow us both to identify optimal assemblages and to dissect how the interplay of pairwise and higher-order interactions shape community-level properties and functions. To our knowledge, the experiments we have just reported represent one of barely a handful of datasets consisting of every potential combination from a library of strains. The fact that it was done in under an hour per replicate gives a sense of the potential our design has to dramatically expand the number of empirical community-function landscapes in the literature.

Discussion

We have introduced a rapid, simple, and inexpensive protocol for assembling all possible combinations of a given set of species using basic laboratory equipment such as a multichannel pipette and 96-well plates. Our protocol enables a single individual to construct all 256 possible consortia from a pool of eight species within an hour, and can be easily expanded to even higher-diversity libraries. Even without the aid of specialized equipment such as robotic liquid handlers, this procedure will make it possible to efficiently generate all communities with up to 10-12 species. Based on our own experience implementing the protocol, we estimate this limit based on manual labor and material constraints rather than conceptual limitations of the protocol itself. Under manual pipetting conditions, this represent a realistic upper bound that can be prepared within a single working day while minimizing handling error and variability.

For laboratories equipped with robotic liquid handlers, the protocol can be easily expanded to accommodate larger libraries, with the main limitation being the amount of plasticware that will be used. The rationale underlying our assembly protocol allows for flexible implementation depending on the specific needs of each laboratory (e.g., it is straightforward to generalize for 384-well plates). Our implementation also has the advantage of keeping pipetting error minimal and independent of consortium size, as we showed by creating all 256 color combinations from eight synthetic colorants. To facilitate automation and adaptation to the needs of other groups, we provide an accompanying R script to facilitate the generation of comprehensive step-by-step protocols for any number of species and pipetting volumes.

In our implementation, the “density homogenization” step ensures that every species is present at a same density across all consortia that it is part of. Note, however, that this homogenization can be bypassed entirely, making it so the density of each species is variable across assemblages (specifically, inversely proportional to the number of strains in a given assemblage). Further variations in initial abundance can be explored, for instance, by treating two (or more) monocultures of a same species at different starting densities as distinct inputs of the protocol – but note that this will quickly increase the total number of assemblages that need to be tested.

We have demonstrated the utility of this protocol in synthetic microbial ecology by generating all possible communities from a collection of eight P. aeruginosa strains and quantifying the strength of pairwise and higher-order functional interactions between them. Our experiment illustrates how the full factorial design of microbial consortia enables for the full characterization of ecological community-function landscapes, which can provide valuable insights into the emergence of community-level functions (such as, but not limited to, optical density at different wavelengths, as we studied here). The entire experiment, from spreading colonies to measuring community function, could be completed from beginning to end by a single experimentalist in less than one working week, and most of that time was taken by waiting for cell growth. In any case, this time frame will also depend on the specific community-level function of interest, and the method used for its quantification. Here we have focused on the absorbance of liquid cell cultures as an illustrative example of a possible readout, but the protocol is compatible with a wide range of other functional, interaction-based, or omics-based assays.

Although initially designed for building combinations of microbial species, our protocol can be readily applied for creating combinations of any soluble (or at least homogeneously distributed) compounds beyond microorganisms. For instance, it could be employed to generate all possible media compositions from a set of resources, all possible combinations of a set of antibiotics, toxins, or even bacteriophages, as well as mixed combinations thereof. We therefore anticipate that our methodology will find extensive applications and utility across diverse fields such as microbiology, ecology, evolution, and biotechnology.